---
title: "Analyzing Predictive Factors of Academic Success in University"
subtitle: "A Linear Regression Analysis of High School Performance, SAT Scores, and Extracurricular Activities"
author: "Student Name"
output: word_document
always_allow_html: true
bibliography: references.bib
---

# Introduction

The challenge of identifying students who will succeed in university settings has long been a critical concern for admission officers. While traditional metrics, such as high school grades and standardized test scores, have been the cornerstone of admission decisions, there is growing recognition that non-academic factors may also play a significant role in predicting academic success. The study looks at the relationship of university academic performance with three potential predictive factors: high school GPA, SAT scores, and extracurricular activity participation. In a sample of 100 fourth-year university students, we carry out linear regression analysis to test these variables in predicting university academic success measured by GPA through the first three years of university study.

# Literature Review

The prediction of academic success in higher education has been extensively studied in educational research. Traditionally, high school GPA has been considered one of the most reliable predictors of college success [@McFarlane2017]. Studies consistently show that high school GPA maintains a strong correlation with college performance, potentially due to its ability to capture both cognitive abilities and non-cognitive factors such as motivation and study habits.

Standardized testing, particularly the SAT, has also been a significant component in admission decisions. Research by the College Board (2019) demonstrates moderate to strong correlations between SAT scores and first-year college GPA. However, some studies suggest that the predictive validity of SAT scores may vary across different student populations and institutional contexts [@beatty2011; @rigdon2009; @kuncel2008].

More recently, attention has turned to the role of extracurricular activities in predicting academic success. Studies by [@marsh2002; @marsh2003] found that participation in extracurricular activities was positively associated with academic achievement, potentially due to the development of time management skills, leadership capabilities, and social capital. However, the relationship between extracurricular involvement and academic performance is complex, with some research suggesting a curvilinear relationship where excessive involvement may detract from academic achievement[@eccles2006]

This study contributes to the existing literature by simultaneously examining these three factors – high school GPA, SAT scores, and extracurricular activities – to determine their relative importance in predicting university academic success. The findings will be particularly valuable for admission officers seeking to develop more comprehensive and evidence-based admission criteria.

# Results

## Exploratory Data Analysis

```{r include=FALSE}
knitr::opts_chunk$set(message = F, warning = F, echo = F)
```

```{r}
require(pacman)
p_load(readxl, tidyverse, tidymodels, readxl, gtsummary, RColorBrewer, ggcorrplot, broom, kableExtra)
```

```{r}
data <- read_xlsx("data.xlsx")
```

### Descriptive Statistics

The sample included 100 fourth-year university students. The university GPA for the first three years was, on average, 7.28 (SD = 2.37) on a 12-point scale. Students in the sample had entered university with a mean high school GPA of 7.98 (SD = 2.05), also on a 12-point scale. The average SAT score was 1,083 (SD = 144) out of a possible 1,600 points. Regarding extracurricular involvement, students reported that they spent an average of 4.8 hrs/wk (SD = 3.3) in organized activities during their last year of high school.

Table 1 presents the descriptive statistics for all variables included in the analysis.

**Table 1**

```{r}
data |> tbl_summary(statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    ))
```

*Note.* GPA = Grade Point Average; SAT = Scholastic Assessment Test. GPA is measured on a 12-point scale. SAT scores range from 400 to 1,600. Extracurricular activities represent average hours per week spent in organized activities during the final year of high school.

### Correlation Analysis

Some notable patterns were observed in the analysis of the relationships between variables. High school GPA was moderately positively related to university GPA, r =.53, p \<.05, indicating that students who did well at high school tended to continue doing well at university. Extracurricular activities were weakly positively related to university GPA, r =.12, and high school GPA, r =.13; neither of these relationships was significant.

Even more intriguingly, SAT scores almost do not correlate with university GPA, r =.02 and a low negative in high school, r = -.10. The result showed that SAT also correlates poorly and a low negative to the extent of extracurricular activities, r = -.17.

```{r}

cor(data) |> 
  ggcorrplot(
    type = "upper",
    method = "circle",
    colors = c("#1A5D1A", "white", "#78C17C"),  # Dark green to white to light green
    lab = TRUE,
    lab_size = 3,
    title = "Correlation Plot of Academic Performance Indicators",
    ggtheme = ggplot2::theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 8)
      )
  ) +
  scale_x_discrete(labels = c("Univ_GPA" = "University GPA",
                             "HS_GPA" = "High School GPA",
                             "SAT" = "SAT Score",
                             "Activities" = "Extracurricular\nActivities")) +
  scale_y_discrete(labels = c("Univ_GPA" = "University GPA",
                             "HS_GPA" = "High School GPA",
                             "SAT" = "SAT Score",
                             "Activities" = "Extracurricular\nActivities"))

```

*Note.* N = 100. GPA = Grade Point Average; SAT = Scholastic Assessment Test.

## Regression Analysis

```{r}
set.seed(123)
data_split <- initial_split(data, prop = 0.8)
train_data <- training(data_split)
test_data <- testing(data_split)

reg_recipe <- recipe(Univ_GPA ~ HS_GPA + SAT + Activities, data = train_data) |>
  step_normalize(all_predictors()) |> 
  step_zv(all_predictors()) |>
  prep()

lm_spec <- linear_reg() |>
  set_engine("lm") |>
  set_mode("regression")

lm_workflow <- workflow() |>
  add_recipe(reg_recipe) |>
  add_model(lm_spec)

lm_fit <- lm_workflow |>
  fit(data = train_data)

model_results <- lm_fit |>
  extract_fit_parsnip() |>
  tidy() |>
  mutate_if(is.numeric, round, 2)

# R-squared and other metrics
model_metrics <- lm_fit |>
  extract_fit_parsnip() |>
  glance() |>
  mutate_if(is.numeric, round, 2)

predictions <- predict(lm_fit, test_data) |>
  bind_cols(test_data)

test_metrics <- metrics(predictions, 
                       truth = Univ_GPA,
                       estimate = .pred) |>
  mutate_if(is.numeric, round, 2)
```

Multiple regression was conducted to predict university GPA based on high school GPA, SAT scores, and extracurricular activities. Overall, the model was significant, F(3, 76) = 11.45, p \<.001, accounting for 17% of the variance in university GPA (R² =.17).

```{r}
model_results |> kbl() |> kable_styling()
```

High school GPA was the sole significant predictor of university GPA, β = 1.27, SE = 0.23, p \<.001. The SAT scores were not a significant predictor, β = 0.16, SE = 0.23, p =.48, neither were the extracurricular activities, β = 0.30, SE = 0.23, p =.20. This model can be represented as:

University GPA = 7.50 + 1.27(HS_GPA) + 0.16(SAT) + 0.30(Activities)

```{r}
model_metrics |> t() |> kbl() |> kable_styling()
```

The model's performance in making predictions was checked against a test set; this yielded an RMSE of 2.15 and an MAE of 1.84. Given these error metrics and considering the actual versus predicted value visualization shown in Figure 1, moderate predictive capability with some deviation from perfect prediction was depicted.

```{r}
test_metrics |> kbl() |> kable_styling()
```

**Figure 1**\

*Scatter Plot of Actual versus Predicted University GPA*

```{r}
# actual vs predicted values
ggplot(predictions, aes(x = Univ_GPA, y = .pred)) +
  geom_point(alpha = 0.5) +
  geom_abline(color = "blue", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Actual vs Predicted University GPA",
    x = "Actual GPA",
    y = "Predicted GPA"
  ) +
  coord_equal()

# Save the model
saveRDS(lm_fit, "final_model.rds")
```

[Figure shows a scatter plot with actual GPA on the x-axis and predicted GPA on the y-axis, with points showing moderate dispersion around the diagonal line of perfect prediction]

Model diagnostics-AIC = 344.99 and BIC = 356.90-and the scatter plot of predicted versus actual values indicate that although the model captures some meaningful relationships, there might be other factors that influence university academic performance not taken into consideration in this model.

These findings indicate that high school GPA is still a valid predictor of success at the university level, but SAT scores and extracurricular activities may be less valid predictors than traditionally believed. Admission officers may want to place more weight on applicants' high school GPAs while de-emphasizing the applicant's SAT scores and extracurricular activities as predictors of academic success.

# Conclusion

This study investigated the predictive relationships between traditional and non-traditional admission criteria and academic success at the university level. The findings yield several important implications for university admission policies and practices.

The results of the analysis revealed that high school GPA is the best predictor of university academic performance, with a significant positive relationship with university GPA. Therefore, our finding is in consonance with earlier studies that academic achievements in high school serve to this day as the gold standard predictor of future success at university. However, academic achievement or success in higher education obviously became an intricate phenomenon that our proposed model could explain no more than 17% of its variation using the variables considered here.

Surprisingly, SAT scores did not show any significant relationship with university GPA, challenging the traditional emphasis placed on standardized testing in admission decisions. This finding adds to the ongoing debate about the value of standardized testing in predicting academic success and supports the growing trend among universities to adopt test-optional policies.

Similarly, the hours spent in extracurricular activities, while showing a positive trend, did not emerge as a statistically significant predictor of university performance. This suggests that while extracurricular involvement may contribute to student development in other important ways, its direct impact on academic performance may be limited or more complex than initially hypothesized.

# References
